---
title: "Midterm 1"
author: "Yumna Moustafa"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 12 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

This exam is due by **12:00p on Thursday, January 28**.  

## Load the tidyverse
If you plan to use any other libraries to complete this assignment then you should load them here.
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
```

## Questions
**1. (2 points) Briefly explain how R, RStudio, and GitHub work together to make work flows in data science transparent and repeatable. What is the advantage of using RMarkdown in this context?**  

#### R is the actual coding language that is being used to perform all actions, and R Studios hosts R and provides a more user-friendly interface that makes it easier to use R, as well as makes creating and viewing plots possible. Github allows users to upload, save, and share their codes on their site, which makes it easier to see past revisions on works, as well as collaborating. RMarkdowns are useful in presenting the code and all plots in a neat way that makes it easier for others to navigate, and makes it possible to view the code and its outputs on github itself, without downloading it. ####

**2. (2 points) What are the three types of `data structures` that we have discussed? Why are we using data frames for BIS 15L?**

#### Data matrices, data frames, and vectors. Data frames are easy to navigate and allow for more operations (from tidyverse especially) to be performed on them, that cannot otherwise be done with a vector or matrix.  ####

In the midterm 1 folder there is a second folder called `data`. Inside the `data` folder, there is a .csv file called `ElephantsMF`. These data are from Phyllis Lee, Stirling University, and are related to Lee, P., et al. (2013), "Enduring consequences of early experiences: 40-year effects on survival and success among African elephants (Loxodonta africana)," Biology Letters, 9: 20130011. [kaggle](https://www.kaggle.com/mostafaelseidy/elephantsmf).  

**3. (2 points) Please load these data as a new object called `elephants`. Use the function(s) of your choice to get an idea of the structure of the data. Be sure to show the class of each variable.**

```{r}
elephants <- readr::read_csv("data/ElephantsMF.csv")
```
```{r}
skim(elephants)
glimpse(elephants)
summary(elephants)
```


**4. (2 points) Change the names of the variables to lower case and change the class of the variable `sex` to a factor.**

```{r}
elephants <- janitor::clean_names(elephants)
elephants$sex <- as.factor(elephants$sex)
```


**5. (2 points) How many male and female elephants are represented in the data?**

```{r}
elephants %>% 
  summarize(femalecount = sum(sex == "F"), malecount = sum(sex == "M"))
```


**6. (2 points) What is the average age all elephants in the data?**

```{r}
mean(elephants$age)
```


**7. (2 points) How does the average age and height of elephants compare by sex?**

```{r}
elephants %>% 
  group_by(sex) %>% 
  summarize(meanage = mean(age), meanheight = mean(height))
```


**8. (2 points) How does the average height of elephants compare by sex for individuals over 25 years old. Include the min and max height as well as the number of individuals in the sample as part of your analysis.**

```{r}
elephants %>% 
  filter(age>25) %>% 
  group_by(sex) %>% 
  summarize(averageheight = mean(height), numbersamples = n(), minheight = min(height), maxheight = max(height))
```


For the next series of questions, we will use data from a study on vertebrate community composition and impacts from defaunation in [Gabon, Africa](https://en.wikipedia.org/wiki/Gabon). One thing to notice is that the data include 24 separate transects. Each transect represents a path through different forest management areas.  

Reference: Koerner SE, Poulsen JR, Blanchard EJ, Okouyi J, Clark CJ. Vertebrate community composition and diversity declines along a defaunation gradient radiating from rural villages in Gabon. _Journal of Applied Ecology_. 2016. This paper, along with a description of the variables is included inside the midterm 1 folder.  

**9. (2 points) Load `IvindoData_DryadVersion.csv` and use the function(s) of your choice to get an idea of the overall structure. Change the variables `HuntCat` and `LandUse` to factors.**

```{r}
ivindodata <- readr::read_csv("data/IvindoData_DryadVersion.csv")
```
```{r}
glimpse(ivindodata)
summary(ivindodata)
skim(ivindodata)
head(ivindodata)

```
```{r}
ivindodata$HuntCat <- as.factor(ivindodata$HuntCat)
ivindodata$LandUse <- as.factor(ivindodata$LandUse)
```


**10. (4 points) For the transects with high and moderate hunting intensity, how does the average diversity of birds and mammals compare?**

```{r}
ivindodata %>% 
  filter(HuntCat == "Moderate" | HuntCat == "High") %>%
  summarize(avedivbirds = mean(Diversity_BirdSpecies), avedivmammals = mean(Diversity_MammalSpecies))
```


**11. (4 points) One of the conclusions in the study is that the relative abundance of animals drops off the closer you get to a village. Let's try to reconstruct this (without the statistics). **

**How does the relative abundance (RA) of apes, birds, elephants, monkeys, rodents, and ungulates compare between sites that are less than 5km from a village to sites that are greater than 20km from a village? The variable `Distance` measures the distance of the transect from the nearest village. Hint: try using the `across` operator.**  

```{r}
greaterthan20dist <- ivindodata %>% 
  filter(Distance>20) %>%
  summarize(across(contains("RA_"), mean)) %>% 
  mutate(Distanceis = "Greater than 20km")
greaterthan20dist

smallerthan5dist <- ivindodata %>% 
  filter(Distance<5) %>% 
  summarize(across(contains("RA_"), mean)) %>% 
  mutate(Distanceis = "Less than 5km")
smallerthan5dist

comparing <- rbind.data.frame(greaterthan20dist, smallerthan5dist)

comparing
  
```

#### The relative abundance of rodents and birds are greater at distances less than 5km from a village, and the relative abundances of apes, elephants, monkeys, and ungulates are greater at distances greater than 20Km from villages ####

**12. (4 points) Based on your interest, do one exploratory analysis on the `gabon` data of your choice. This analysis needs to include a minimum of two functions in `dplyr.`**

#### Bird diversity is greater in lands used as parks, but the relative abundance of birds is greater where the land is used for logging ####

```{r}
ivindodata %>% 
  filter(LandUse != "Neither") %>% 
  group_by(LandUse) %>%
  summarise(birddiversity = mean(Diversity_BirdSpecies), birdabundance = mean(RA_Birds))
```




